---
title: "Spatial Analysis of Environmental Factors Related to Large Tornado Groups"
author: "Zoe Shroder"
date: "7/31/2018"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

Research to be submitted to the **Electronic Journal of Severe Storms Meteorology**

Q: What are the broad geographic variations in CAPE and shear associated with tornado groups? How widely do these environmental characteristics vary geographically? Q: Which environmental variable (CAPE or shear) best explains group-level average tornado energy?

Why is this important? Tornado groups represent a concentrated large-scale risk for damage and casualties. In fact 3/4 of all fatalities occur on days with the most tornadoes within an outbreak.

This analysis is done at the scale of tornado groups rather than at local scales. 

To minimize controversy related to terminology tornado 'group' is used to refer to tornadoes occurring together in space and time. The word 'outbreak' has a specific definition that is not strictly applied here. 

Analysis and modeling makes use of the Storm Prediction Center's tornado data set.

## Part 1: Tornado data

Set working directory and load packages.
```{r}
library(tidyverse)
library(lubridate)
library(sf)
library(tmap)
library(USAboundaries)
library(rgeos)
```

The newest GIS shapefile contains missing geometries for more than 30% of the tornadoes. The number of missing geometries is highest after 1995. Instead use the csv file from https://www.spc.noaa.gov/wcm/#data  Use the start lon/lat and create a `sp` object then convert to `sf`. Set the coordinate reference system (crs) to ESPG 4326.
```{r}
Tor.df <- read.csv(file = "1950-2017_actual_tornadoes.csv")
Tor.spdf <- Tor.df
rm(Tor.df)
sp::coordinates(Tor.spdf) <- ~ slon + slat
Tor.sfdf <- st_as_sf(Tor.spdf)
st_crs(Tor.sfdf) <- 4326
```

Remove tornadoes in Hawaii, Alaska, and Puerto Rico and those occurring before 1994. For missing EF ratings use https://www.spc.noaa.gov/wcm/OneTor_F-scale-modifications.pdf
```{r}
Tor.sfdf <- Tor.sfdf %>%
  filter(yr >= 1994, #start in 1994 because of comprehensive radar
         !st %in% c("AK", "PR", "HI")) %>% #remove tornadoes occurring in AK, PR, and HI
  mutate(mag = ifelse(mag == -9 & len <= 5, 0, mag),
         mag = ifelse(mag == -9 & len > 5, 1, mag))
```

Add a data/time column also add columns for path length, width, and area in metric units. Leave the time zone as native CDT. Create a convective day (6AM to 6AM) column taking hours 00:00:00 -> 05:59:59 and assigning it to the previous date (this associates the previous day's date to tornadoes occurring up to 6 hours after local midnight).
```{r}
Tor.sfdf <- Tor.sfdf %>%
  mutate(dy = format(as.Date(date,format="%m/%d/%y"), "%d"),
         DateTime = as.POSIXct(paste(yr, mo, dy, time), format = "%Y%m%d%H:%M:%S"),
         Hour = hour(DateTime),
         Year = year(DateTime),
         cDateTime = DateTime - as.difftime(6, unit = "hours"),
         cDate = as.Date(as_datetime(ifelse(Hour < 6, (DateTime - 86400), cDateTime), tz = Sys.timezone())),
         Length = len * 1609.34,
         Length = ifelse(Length == 0, min(Length[Length > 0]), Length), #takes care of zero length
         Width = wid * .9144,
         Width = ifelse(Width == 0, min(Width[Width > 0]), Width), #takes care of zero width
         Width = ifelse(Year >= 1995, Width * pi/4, Width), #takes care of change: avg to max
         cas = inj + fat,
         AreaPath = Length * Width,
         Ma = factor(month.abb[mo], levels = month.abb[1:12])) %>%
  sf::st_sf()
max(Tor.sfdf$yr)
```

The geometry type is `POINT`. Each tornado is a single location (start location). 

Add energy dissipation per tornado. Use the empirical model for tornado winds by EF rating taken from Table 3-1 of NRC 2007. Percent area by EF rating for each EF category. Threshold wind speeds (m/s) are a lower bound 3 sec gusts on the operational EF Scale (Table 2-1 of NRC2007). This is based on work by Fricker et al.

$E = A_p \rho \sum_{j=0}^{J} w_j v_j^{3},$ where $A_p$ is the area of the path, $\rho$ is area density [1 kg/m^3]  $v_j$ is the midpoint wind speed for each rating, and $w_j$ is the corresponding fraction of path area by EF rating. With no upper bound on the EF5 wind speeds, the midpoint wind speed is set at 97 m~s$^{-1}$ (7.5 m~s$^{-1}$ above the threshold wind speed consistent with the EF4 midpoint speed relative to its threshold)
```{r}
perc <- c(1, 0, 0, 0, 0, 0, 
         .772, .228, 0, 0, 0, 0,
         .616, .268, .115, 0, 0, 0,
         .529, .271, .133, .067, 0, 0,
         .543, .238, .131, .056, .032, 0,
         .538, .223, .119, .07, .033, .017)
percM <- matrix(perc, ncol = 6, byrow = TRUE)
threshW <- c(29.06, 38.45, 49.62, 60.8, 74.21, 89.41)
midptW <- c(diff(threshW)/2 + threshW[-length(threshW)], threshW[length(threshW)] + 7.5)
ef <- Tor.sfdf$mag + 1
EW3 <- numeric()
for(i in 1:length(ef)) EW3[i] = midptW^3 %*% percM[ef[i], ]
Tor.sfdf <- Tor.sfdf %>%
  mutate(ED = EW3 * AreaPath)
```

Determine the distance between tornadoes in space and time. Use a projection not lat/lon. See https://epsg.io/102004. Extract the coordinates of the start locations as a N by 2 matrix of numbers, where N is the number of tornadoes. Also extract the date-time as a vector of class `POSIXct`.
```{r}
Tor.sfdfT <- st_transform(Tor.sfdf, crs = 102004)
space <- st_coordinates(Tor.sfdfT)
time <- Tor.sfdf$DateTime
```

Next compute pairwise Euclidean distances in space and, separately, in time using the `dist()` function. Divide the spatial distance by 10 so that the values are commensurate with the time 'distance' based on the assumption of 10 meters every second for an average speed of tornado-generating storms. 

Compare: Distance from New York to Denver is 2.622 x 10^6 meters. There are 3.154 x 10^7 seconds in a year. This will capture the historic multiday tornado outbreaks. For analysis we want to consider each day in the multiday group separately. As the value of the divisor increases cluster areas get larger.
```{r}
ds <- dist(space) / 10
dt <- dist(time)
dst <- ds + dt
rm(ds, dt) # helps control memory allocation
```

Distances are saved as an object of class `dist` containing a vector of length N * (N-1)/2, which is the number of unique point pairs.

Next group the tornadoes based on the space-time distances. This is done with the `hclust()` (hierarchical cluster) function. Initially, each tornado is assigned to its own group and then the algorithm joins the two closest tornadoes determined by values in `dst`. The algorithm continues by joining tornadoes (and tornado groups) until there is a single large group.

The single linkage method (`method = "single"`) is related to the minimal spanning tree (MST) and adopts a 'friends of friends' grouping strategy. An edge-weighted graph is a graph where each edge has a weight (or cost). Here weights are space-time distances between tornadoes. A MST of an edge-weighted graph is a spanning tree whose weight (the sum of the weights of its edges) is no larger than the weight of any other spanning tree. A spanning tree of a graph on N vertices (tornado centroids) is a subset of N-1 edges that form a tree (Skiena 1990, p. 227).

The `cutree()` function is used to extract a group number for each tornado. Tornadoes in each group are close in space & time. Here the tree is cut at a height of 100000 space-time units. Making `h` smaller results in smaller groups (fewer tornadoes per group).
```{r}
stime <- proc.time()
tree <- hclust(dst, method = "single")
groupNumber <- as.integer(cutree(tree, h = 100000))
proc.time() - stime
```

 user  system elapsed 
 21.256   1.677  22.983 

Add the group number to each tornado. 
```{r}
Tor.sfdfT$groupNumber <- groupNumber
```

Compute group-level statistics. Keep only tornado groups with at least 30 tornadoes.
```{r}
Groups.sfdfT <- Tor.sfdfT %>%
  group_by(groupNumber) %>%
  summarize(Year = first(Year),
            Month = first(mo),
            FirstDate = first(date),
            LastDate = last(date),
            Name = paste(FirstDate, "to", LastDate),
            FirstcDate = first(cDate),
            LastcDate = last(cDate),
            ncD = n_distinct(cDate),
            nT = n(),
            n0 = sum(mag == 0),
            n1 = sum(mag == 1),
            n2 = sum(mag == 2),
            n3 = sum(mag == 3),
            n4 = sum(mag == 4),
            n5 = sum(mag == 5),
            GroupTotalED = sum(ED),
            Name2 = paste(round(GroupTotalED/10^12), "TW"),
            maxEF = max(mag),
            nD = n_distinct(date),
            StartTime = first(DateTime),
            EndTime = last(DateTime),
            Duration = difftime(EndTime, StartTime, units = "secs")) %>%
  filter(nT >= 30)
dim(Groups.sfdfT)
```

Over the period 1994-2017 there are 198 tornado groups with at least 30 tornadoes.

Make some tables: Top 10 groups by number of tornadoes, energy dissipation, and duration.
```{r}
Groups.sfdfT %>%
  top_n(n = 10, wt = nT) %>%
  arrange(desc(nT))

Groups.sfdfT %>%
  top_n(n = 10, wt = GroupTotalED) %>%
  arrange(desc(GroupTotalED))

Groups.sfdfT %>%
  top_n(n = 10, wt = Duration) %>%
  arrange(desc(Duration))

Groups.sfdfT %>%
  top_n(n = 10, wt = nD) %>%
  arrange(desc(nD))
```

Map the locations (start points) of the top 12 tornado groups by energy dissipation. First select the top 12 groups by total ED. Then create a simple features data frame using only the tornadoes in these groups by filtering on group number.
```{r}
df <- as.data.frame(Groups.sfdfT) %>%
  top_n(n = 12, wt = GroupTotalED)

sfdf <- Tor.sfdfT %>%
   filter(groupNumber %in% df$groupNumber) %>%
   mutate(EF = factor(mag))

sfdf2 <- left_join(sfdf, df, by = "groupNumber") %>%
  arrange(desc(GroupTotalED))
```

Get state borders and use the `tm_shape()` function.
```{r}
sts <- state.name[!state.name %in% c("Alaska", "Hawaii")]
stateBorders <- us_states(states = sts)

tm_shape(sfdf2) +
  tm_facets(by = "Name", ncol = 3)  +
  tm_symbols(col = "EF", palette = "Reds", n = 6) +
  tm_credits(text = df$Name2, position = c("left", "bottom"), bg.color = "white") +
tm_shape(stateBorders) + 
  tm_borders(col = "grey") +
#  tm_compass() + tm_scale_bar() +
  tm_layout(legend.outside = TRUE)
```

Distribution of the number of days.
```{r}
dfnD <- as.data.frame(Groups.sfdfT) %>%
  arrange(desc(nD))
table(dfnD$nD)
```

Frequency by month of the start day of multi-day groups of at least 4, 5, and 6 days in length.
```{r}
as.data.frame(Groups.sfdfT) %>%
  filter(nD >= 4) %>%
  count(Month) %>%
  mutate(prop = prop.table(n))

as.data.frame(Groups.sfdfT) %>%
  count(Month, nD) %>%
  mutate(prop = prop.table(n))
```

Filter on tornadoes to remove those not part of a large group. Group by group number and convective dates. Remove group days with fewer than 10 tornadoes.
```{r}
BigDaysInLargeGroups.sfdfT <- Tor.sfdfT %>%
  filter(groupNumber %in% Groups.sfdfT$groupNumber) %>%
  group_by(groupNumber, cDate) %>%
  summarize(nT = n(),
            GroupDayTotalED = sum(ED),
            GroupDayMaxED = max(ED),
            GroupDayMeanED = mean(ED),
            GroupDayCas = sum(cas),
            GroupDayFat = sum(fat)) %>%
  filter(nT >= 10) %>%
  mutate(Year = year(cDate),
         Mo = month(cDate),
         Month = format(cDate, "%m"), # this is needed to preserve the leading zeros
         Day = format(cDate, "%d"))
dim(BigDaysInLargeGroups.sfdfT)
```

There are 410 big days in large groups. These will be used for further analysis and modeling.

Pecentage of all casualties and fatalities occurring on big days in large outbreaks.
```{r}
sum(BigDaysInLargeGroups.sfdfT$GroupDayCas)/sum(Tor.sfdfT$cas) * 100
sum(BigDaysInLargeGroups.sfdfT$GroupDayFat)/sum(Tor.sfdfT$fat) * 100
```

Table the number of days per group. Also look at the day counts by year and month.
```{r}
table(table(BigDaysInLargeGroups.sfdfT$groupNumber))
table(BigDaysInLargeGroups.sfdfT$Year)
table(BigDaysInLargeGroups.sfdfT$Month)
```

71% of all big days in large groups occur during April, May, and June.

Top 10 biggest tornado days within the largest groups.
```{r}
BigDaysInLargeGroups.sfdfT %>%
  top_n(n = 10, nT) %>%
  arrange(desc(nT))
```

Extract the geographic centroids of tornado genesis on big tornado days by creating an `sf` object using `st_centroid()` that reduces the MULTIPOINT geometry to a POINT geometry. Compute the area of the group and the number of tornadoes per area (density).
```{r}
groupDayCentroids.sfdfT <- st_centroid(BigDaysInLargeGroups.sfdfT)
groupDayCentroids.sfdfT$groupArea <- st_area(st_convex_hull(BigDaysInLargeGroups.sfdfT))
groupDayCentroids.sfdfT$groupDensity <- groupDayCentroids.sfdfT$nT/groupDayCentroids.sfdfT$groupArea
```

Plot the centroid for each group day and size the aesthetic by the number of tornadoes.
```{r}
tm_shape(groupDayCentroids.sfdfT) +
  tm_symbols(size = "nT", title.size = "Count", legend.size.is.portrait = TRUE) + 
tm_shape(stateBorders) + 
  tm_borders() +
  tm_compass() + tm_scale_bar() +
  tm_layout(legend.bg.color = "white", legend.text.size = .75)
```

Tornado group day centers do not have an obvious population bias.

Time series graphs
```{r}
groupDayCentroids.sfdfT %>%
  group_by(Year) %>%
  summarize(avgArea = mean(groupArea)) %>%
  ggplot(., aes(x = Year, y = as.numeric(avgArea) / 10^10)) +
    scale_y_continuous(limits = c(0, NA)) +
    geom_point() +
    geom_smooth(method = lm) +
    ylab("Group Area [100 x 100 sq. km]") +
    theme_minimal()

groupDayCentroids.sfdfT %>%
  group_by(Year) %>%
  summarize(avgDensity = mean(groupDensity)) %>%
  ggplot(., aes(x = Year, y = as.numeric(avgDensity) * 10^10)) +
    scale_y_continuous(limits = c(0, NA)) +
    geom_point() +
    geom_smooth(method = lm) +
    ylab("Group Density [Tornadoes/100 x 100 sq. km]") +
    theme_minimal()
```

Obtain the group day hulls. Transform the CRS to match that of the environmental data raster grids.
```{r}
BigDaysInLargeGroups.sfdfT<-as.data.frame(BigDaysInLargeGroups.sfdfT)
BigDaysInLargeGroups.sfdfT<-st_sf(BigDaysInLargeGroups.sfdfT)

BigDaysInLargeGroups.sfdfT <- st_convex_hull(BigDaysInLargeGroups.sfdfT)
BigDaysInLargeGroups.sfdfT <- st_transform(BigDaysInLargeGroups.sfdfT, 
  crs = "+proj=lcc +lat_1=50 +lat_2=50 +lat_0=50 +lon_0=-107 +x_0=0 +y_0=0 +a=6371200 +b=6371200 +units=m +no_defs")

tm_shape(BigDaysInLargeGroups.sfdfT) +
  tm_polygons(alpha = .1) + 
tm_shape(stateBorders, is.master = TRUE) + 
  tm_borders() +
  tm_compass() + tm_scale_bar() 
```

Obtain the state and county boundaries from the **USAboundaries** package. 
```{r}
library(USAboundaries)
states.sf <- us_states()
counties.sf <- us_counties()
```

Transform the CRS of the county boundaries to match the CRS of the BigDaysInLargeGroups.sfdfT
```{r}
counties.sf <- st_transform(counties.sf, 
                            crs = st_crs(BigDaysInLargeGroups.sfdfT))
```

Intersect
```{r}
gI <- st_intersects(BigDaysInLargeGroups.sfdfT, counties.sf, sparse = FALSE)
#gI <- st_intersects(BigDaysInLargeGroups.sfdfT, counties.sf)
colnames(gI) <- counties.sf$name
counties.sf$GroupDayCount <- colSums(gI)
```

Remove Alaska, Hawaii, and Puerto Rico from the `states.sf` data frame.
```{r}
states.sf <- states.sf %>% 
  filter(stusps != "AK",
         stusps != "PR",
         stusps != "HI")
```

Create a map of the outbreak density in each county in the United States. Fill by the `GroupDayCount` column. 
```{r}
tm_shape(states.sf, projection = "laea_NA") +
  tm_borders() +
tm_shape(counties.sf) +
   tm_fill("GroupDayCount",
            title = "Count",
            palette = "Reds", n = 5) +
  tm_borders(col = "gray", alpha = .3) +
  tm_compass() + 
  tm_scale_bar() +
  tm_format_Europe(legend.position = c("right", "bottom"),
                   attr.position = c("left", "bottom"),
                   legend.frame = FALSE,
                   title = "Tornado Group Days by County [1994-2017]",
                   title.size = 1.3,
                   title.position = c("left", "TOP"),
                   inner.margins = c(.05, .05, .1, .05))
```

## Part 2: Environmental data

Get environmental data at 18Z (2p local) on the convective day. Also try this site https://rda.ucar.edu/datasets/ds608.0/ Set up a vector of URLs as character strings. Data are not available after October 2014.
```{r, eval=FALSE}
library(lubridate)
df <- as.data.frame(BigDaysInLargeGroups.sfdfT) %>%
  filter(Year <= 2014 & Mo < 10) %>%
  mutate(Yr = Year,
         YrMo = paste0(Year, Month),
         YrMoDa = paste0(YrMo, Day),
         slug2 = paste0(YrMo, "/", YrMoDa, "/", "narr-a_221_", YrMoDa, "_1800_000.grb"),
         slug = paste0("https://nomads.ncdc.noaa.gov/data/narr/", slug2)) %>%
  dplyr::select(slug)
slug <- df$slug
```

Download the grib files. Note: for the first run we only get the first 50 days.
```{r, eval=FALSE}
for(i in 1:50){
    download.file(slug[i], paste0("NARRdata", i, ".grb"), mode = "wb")
    }
```

Read the grib files as raster bricks and assign the CAPE and helicity variables to separate raster layers. Extract the average CAPE and helicity within each of the big days in large groups hulls.
```{r, eval=FALSE}
library(raster)
aCAPE <- numeric()
aHLCY <- numeric()
aCIN <- numeric()

for(i in 1:50){
  print(i)
  rb <- brick(paste0("NARRdata", i, ".grb"))
  CAPE.rl <- raster(rb, layer = 375)
  HLCY.rl <- raster(rb, layer = 323)
  CIN.rl <- raster(rb, layer = 376)
  aCAPE <- c(aCAPE, mean(raster::extract(CAPE.rl, BigDaysInLargeGroups.sfdfT[i, ], fun = mean)))
  aHLCY <- c(aHLCY, mean(raster::extract(HLCY.rl, BigDaysInLargeGroups.sfdfT[i, ], fun = mean)))
  aCIN <- c(aCIN, mean(raster::extract(CIN.rl, BigDaysInLargeGroups.sfdfT[i, ], fun = mean)))
}
```

Add environmental data values to the group day means data frame.
```{r, eval=FALSE}
sfdf <- BigDaysInLargeGroups.sfdfT[1:50, ]
sfdf$aCAPE <- aCAPE
sfdf$aHLCY <- aHLCY
sfdf$aCIN <- aCIN
```

Random effects.
```{r, eval=FALSE}
sfdf %>%
   group_by(Mo) %>%
   summarize(mED = mean(GroupDayTotalED)/10^12,
             nT = sum(nT))
```

Models
```{r, eval=FALSE}
library(lme4)
model0 <- lmer(log(GroupDayMeanED) ~ scale(aHLCY) + scale(aCAPE) * scale(aCIN) + (1|Mo), data = sfdf)
summary(model0)

model0a <- lmer(log(GroupDayMeanED) ~ scale(aHLCY) + scale(aCAPE) * scale(aCIN) + (1|Mo), weights = nT, data = sfdf)
summary(model0a)

model1 <- lmer(log(GroupDayTotalED) ~ scale(aCAPE) * scale(aCIN) + (scale(aHLCY)|Mo), data = sfdf)
summary(model1)

model2 <- lmer(log(GroupDayMaxED) ~  scale(aHLCY) + scale(aCAPE) * scale(aCIN) + (1|Mo), data = sfdf)
summary(model2)

model3 <- lmer(log(GroupDayCas) ~ scale(aHLCY) + scale(aCAPE) * scale(aCIN) + (1|Mo), data = sfdf)
summary(model3)
```
